---
title: "Kernel Architecture Analysis: GA104 Tensor Core GEMM"
subtitle: "Empirical Performance Models for Hand-Optimized CUDA/SASS on RTX 3070 Ti"
author: "Philipp Thoss"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    self-contained: true
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
engine: knitr
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

theme_gpu <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey40", size = 11),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
theme_set(theme_gpu)

gpu_green  <- "#76B900"
gpu_blue   <- "#0071C5"
gpu_red    <- "#E74C3C"
gpu_orange <- "#F39C12"
gpu_purple <- "#8E44AD"
gpu_grey   <- "#95A5A6"
```

## Introduction

This document presents a quantitative analysis of hand-optimized CUDA/SASS
kernels targeting the NVIDIA RTX 3070 Ti (GA104, sm_86, Ampere architecture).
All kernels are written from scratch --- no cuBLAS, cuDNN, or PyTorch --- and
compiled with `nvcc -arch=sm_86 -O2`. SASS-level inspection and hand-editing
via CuAssembler provides ground-truth for hardware behavior.

The analysis covers 23 empirical insights across 5 development phases,
distilled into quantitative models of the three resources that govern
performance on GA104: **registers**, **shared memory**, and **warp occupancy**.

### Hardware Specifications

| Resource | GA104 (RTX 3070 Ti Laptop) |
|---|---|
| Streaming Multiprocessors | 48 |
| CUDA cores / SM | 128 |
| Tensor Cores / SM | 4 (3rd gen) |
| Registers / SM | 65,536 (32-bit) |
| Shared memory / SM | 100 KB max configurable |
| L2 cache | 4 MB |
| DRAM bandwidth | 608 GB/s |
| FP32 peak (FFMA) | 21.7 TFLOPS |
| FP16 Tensor peak (HMMA) | 174 TFLOPS |
| INT8 Tensor peak (IMMA) | 696 TOPS |

## The Occupancy Model {#sec-occupancy}

Performance on GA104 is dominated by **warp occupancy** --- the number of
active warps available to hide DRAM latency (~300 cycles). The occupancy
is jointly determined by three per-block resources: thread count,
register usage, and shared memory.

```{r fig-occupancy-model}
#| fig-cap: "GA104 occupancy model: achievable blocks per SM as a function of registers per thread and shared memory per block. The 50 KB cliff (dashed red) halves occupancy. Kernel variants are plotted at their measured resource usage."

expand_grid <- function(...) tidyr::crossing(...)

occupancy <- expand_grid(
  regs = seq(32, 255, by = 1),
  smem_kb = seq(0, 100, by = 1)
) |>
  mutate(
    max_by_regs = pmin(floor(65536 / (regs * 256)), 4),
    max_by_smem = ifelse(smem_kb == 0, 4, pmin(floor(100 / smem_kb), 4)),
    blocks_per_sm = pmin(max_by_regs, max_by_smem),
    warps_per_sm = blocks_per_sm * 8
  )

kernels <- tibble::tribble(
  ~label,                ~regs, ~smem_kb, ~gflops, ~variant,
  "HGEMM naive",           40,        0,    7849, "HGEMM",
  "HGEMM tiled (8w)",     148,       40,   16687, "HGEMM",
  "HGEMM 16-warp",         64,       32,   18700, "HGEMM",
  "Online-quant v2",      239,       48,   12707, "Online-quant",
  "In-place",             213,       40,   16646, "Online-quant",
  "Bank-free",            211,       40,   17070, "Online-quant",
  "Warp-specialized",     255,       40,   16039, "Online-quant",
  "IGEMM 8w 128x128",     120,       24,   24533, "IGEMM (INT8)",
  "IGEMM 8w 128x256",     210,       24,   27591, "IGEMM (INT8)",
  "IGEMM cp.async 64x64",  80,        8,   20688, "IGEMM (INT8)"
) |>
  mutate(
    blocks = pmin(
      pmin(floor(65536 / (regs * 256)), 4),
      ifelse(smem_kb == 0, 4, pmin(floor(100 / smem_kb), 4))
    ),
    warps = blocks * 8
  )

# Use 256 threads per block for the heatmap (standard for 8-warp blocks)
occupancy_256 <- expand_grid(
  regs = seq(32, 255, by = 1),
  smem_kb = seq(1, 100, by = 1)
) |>
  mutate(
    max_by_regs_256 = pmin(floor(65536 / (regs * 256)), 4),
    max_by_smem = pmin(floor(100 / smem_kb), 4),
    blocks = pmin(max_by_regs_256, max_by_smem),
    warps = blocks * 8
  )

ggplot(occupancy_256, aes(x = regs, y = smem_kb, fill = factor(blocks))) +
  geom_tile() +
  scale_fill_manual(
    values = c("0" = "#2C3E50", "1" = gpu_red, "2" = gpu_orange,
               "3" = gpu_blue, "4" = gpu_green),
    name = "Blocks / SM"
  ) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = 240, y = 53, label = "50 KB cliff",
           color = "red", fontface = "bold", size = 3.5) +
  geom_vline(xintercept = 128, linetype = "dotted", color = "white", linewidth = 0.5) +
  annotate("text", x = 131, y = 95, label = "128 reg\n(2 blk limit)",
           color = "white", size = 2.8, hjust = 0) +
  geom_point(data = kernels, aes(x = regs, y = smem_kb),
             shape = 21, size = 3, fill = "white", color = "black", stroke = 1.2) +
  geom_text(data = kernels, aes(x = regs, y = smem_kb, label = label),
            size = 2.5, hjust = -0.1, vjust = -0.8, color = "white", fontface = "bold") +
  labs(
    title = "GA104 Occupancy Landscape",
    subtitle = "Blocks per SM as f(registers/thread, shared memory/block) at 256 threads/block",
    x = "Registers per thread",
    y = "Shared memory per block (KB)"
  ) +
  coord_cartesian(xlim = c(32, 260), ylim = c(0, 100)) +
  theme(legend.position = "right")
```

@fig-occupancy-model reveals the two critical thresholds on GA104:

1. **The 50 KB shared memory cliff**: Any block using >50 KB smem is limited to
   1 block/SM. This halves warp occupancy from 8 to 4, exposing ~300-cycle DRAM
   stalls.

2. **The 128-register boundary**: At 256 threads/block, >128 registers/thread
   exceeds the 65,536 register budget for 2 blocks/SM.

The 16-warp HGEMM kernel (64 regs, 32 KB) sits in the 2-blocks/SM sweet spot,
achieving 32 warps/SM --- ideal latency hiding. The online-quant variants
(211--255 regs, 40 KB) are locked at 1 block/SM.


## Performance Progression {#sec-progression}

```{r fig-igemm-progression}
#| fig-cap: "INT8 IGEMM throughput evolution across optimization stages. Each bar represents a distinct kernel variant at M=N=K=4096. The dominant optimization (cp.async pipelining, +35%) exceeds all prior gains combined."

igemm_data <- tibble::tribble(
  ~kernel,                    ~tops,  ~stage,         ~order,
  "Naive (global)",           10897,  "Baseline",         1,
  "Tiled 64x64",              15078,  "Tiling",           2,
  "Hand-tuned S02",           15341,  "SASS tuning",      3,
  "Pipelined LDG",            18054,  "Pipelining",       4,
  "Pipelined cp.async",       20688,  "Pipelining",       5,
  "8-warp 128x128",           24533,  "Occupancy",        6,
  "8-warp 128x256",           27591,  "Occupancy",        7,
  "Persistent 128x256",       27383,  "Persistent",       8
) |>
  mutate(
    kernel = factor(kernel, levels = kernel[order(order)]),
    pct_peak = tops / 696000 * 100,
    vs_naive = tops / 10897
  )

stage_colors <- c(
  "Baseline" = gpu_grey,
  "Tiling" = gpu_blue,
  "SASS tuning" = gpu_purple,
  "Pipelining" = gpu_green,
  "Occupancy" = gpu_orange,
  "Persistent" = gpu_red
)

ggplot(igemm_data, aes(x = kernel, y = tops / 1000, fill = stage)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1fk", tops / 1000)),
            vjust = -0.3, size = 3.2) +
  scale_fill_manual(values = stage_colors, name = "Optimization stage") +
  scale_y_continuous(
    labels = label_comma(suffix = "k"),
    sec.axis = sec_axis(~ . / 696 * 100, name = "% of INT8 peak (696 TOPS)")
  ) +
  labs(
    title = "INT8 IGEMM Optimization Progression",
    subtitle = "M = N = K = 4096, RTX 3070 Ti (GA104, sm_86)",
    x = NULL, y = "Throughput (TOPS)"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 10))
```

The progression in @fig-igemm-progression reveals a consistent pattern: **algorithmic
restructuring** (tiling, pipelining) delivers 10--35% gains, while **micro-level tuning**
(SASS stall codes) delivers 1--2%. The single largest jump (+35%) comes from
software pipelining with `cp.async`, which decouples memory loading from the
warp instruction stream.


## cp.async Benefit Model {#sec-cpasync}

The most surprising finding across the project: `cp.async` (LDGSTS in SASS) is
not universally beneficial. Its impact depends on the **compute-to-load ratio**
per tile --- the number of Tensor Core instructions executed between consecutive
shared memory loads.

```{r fig-cpasync-benefit}
#| fig-cap: "cp.async speedup vs compute density per tile. Benefits scale inversely with compute/load ratio: short compute phases (IGEMM, 8 IMMA/tile) gain +35%, while long compute phases (Flash Attention, 64 HMMA/tile) lose 4--5%. The crossover occurs near 16--32 instructions per tile."

cpasync_data <- tibble::tribble(
  ~kernel,          ~tc_per_tile, ~speedup, ~type,
  "IGEMM 64x64\n(BK=32)",      8,   1.35,  "IMMA (INT8)",
  "IGEMM 64x64\n(BK=64)",     16,   0.946, "IMMA (INT8)",
  "Cross-attn\n(Bc=64)",      64,   0.84,  "HMMA (FP16)",
  "Self-attn\n(Bc=64)",       64,   0.95,  "HMMA (FP16)"
)

# Model: benefit ~ 1 / sqrt(tc_per_tile)
model_x <- seq(4, 80, by = 1)
model_y <- 1 + 0.35 * (8 / model_x)^0.7

ggplot(cpasync_data, aes(x = tc_per_tile, y = (speedup - 1) * 100)) +
  geom_line(data = tibble(x = model_x, y = (model_y - 1) * 100),
            aes(x = x, y = y), color = gpu_grey, linewidth = 0.8, linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "solid", color = "grey50") +
  geom_point(aes(color = type, shape = type), size = 4.5) +
  geom_text(aes(label = kernel), vjust = -1.2, size = 3.2) +
  scale_color_manual(values = c("IMMA (INT8)" = gpu_green, "HMMA (FP16)" = gpu_blue)) +
  scale_shape_manual(values = c("IMMA (INT8)" = 16, "HMMA (FP16)" = 17)) +
  annotate("rect", xmin = 0, xmax = 12, ymin = -10, ymax = 40,
           alpha = 0.08, fill = gpu_green) +
  annotate("text", x = 6, y = 37, label = "cp.async\nhelpful", color = gpu_green,
           fontface = "bold", size = 3.5) +
  annotate("rect", xmin = 30, xmax = 80, ymin = -10, ymax = 40,
           alpha = 0.08, fill = gpu_red) +
  annotate("text", x = 55, y = 37, label = "cp.async\nharmful", color = gpu_red,
           fontface = "bold", size = 3.5) +
  labs(
    title = "cp.async Benefit vs Compute Density",
    subtitle = "At 8 warps/SM: benefit scales inversely with Tensor Core instructions per tile",
    x = "Tensor Core instructions per tile",
    y = "Speedup from cp.async (%)",
    color = "TC type", shape = "TC type"
  ) +
  coord_cartesian(xlim = c(0, 75), ylim = c(-10, 40))
```

The model in @fig-cpasync-benefit explains a key design rule:

> **At 8 warps/SM, cp.async helps only when the compute phase is shorter than
> the DRAM latency window (~300 cycles / 8 warps = ~37 cycles per warp).**

With 8 IMMA instructions per tile (each ~4 cycles = 32 cycles total), the compute
phase is shorter than the per-warp latency window, leaving idle cycles that
cp.async fills. With 64 HMMA instructions per tile (each ~8 cycles = 512 cycles total),
the compute phase alone exceeds the DRAM latency, and warp interleaving already
provides full coverage.


## Register Budget Economics {#sec-registers}

Registers are the scarcest resource per thread. On GA104, the 65,536-register
budget per SM creates hard occupancy boundaries. The WMMA API's fragment-based
programming model makes register allocation transparent.

```{r fig-register-budget}
#| fig-cap: "Register budget breakdown for key kernel variants. The accumulator dominates allocation in all cases. The 16-warp HGEMM achieves 64 registers by halving the warp tile from 2x4 to 2x2, cutting accumulator cost from 64 to 32 registers."

reg_data <- tibble::tribble(
  ~kernel,              ~accumulator, ~a_frag, ~b_frag, ~control, ~total, ~blocks,
  "HGEMM tiled\n(8w, 2x4)",    64,       16,      32,      36,     148,    1,
  "HGEMM 16-warp\n(16w, 2x2)", 32,       16,      16,       0,      64,    2,
  "Online-quant\nbankfree",    128,       16,      32,      35,     211,    1,
  "IGEMM 8w\n128x256",         128,       16,      64,       2,     210,    1
) |>
  mutate(kernel = factor(kernel, levels = kernel))

reg_long <- reg_data |>
  select(kernel, accumulator, a_frag, b_frag, control) |>
  pivot_longer(-kernel, names_to = "component", values_to = "regs") |>
  mutate(component = factor(component,
    levels = c("control", "b_frag", "a_frag", "accumulator"),
    labels = c("Control + pointers", "B fragments", "A fragments", "Accumulator")))

ggplot(reg_long, aes(x = kernel, y = regs, fill = component)) +
  geom_col(width = 0.65) +
  geom_hline(yintercept = 128, linetype = "dashed", color = gpu_red, linewidth = 0.8) +
  annotate("text", x = 4.4, y = 134, label = "128 reg limit\n(2 blocks/SM)",
           color = gpu_red, fontface = "bold", size = 3, hjust = 1) +
  geom_hline(yintercept = 255, linetype = "dotted", color = "grey50") +
  annotate("text", x = 4.4, y = 248, label = "255 max",
           color = "grey50", size = 3, hjust = 1) +
  geom_text(data = reg_data, aes(x = kernel, y = total, label = paste0(total, " regs")),
            vjust = -0.4, size = 3.5, fontface = "bold", inherit.aes = FALSE) +
  scale_fill_manual(
    values = c("Accumulator" = gpu_orange, "A fragments" = gpu_blue,
               "B fragments" = gpu_green, "Control + pointers" = gpu_grey),
    name = "Component"
  ) +
  labs(
    title = "Register Budget Breakdown",
    subtitle = "WMMA fragment registers dominate allocation; accumulator size determines occupancy tier",
    x = NULL, y = "Registers per thread"
  ) +
  coord_cartesian(ylim = c(0, 270))
```

@fig-register-budget reveals the central trade-off: **the accumulator fragment count
determines the occupancy tier**. Online-quant requires both INT32 WMMA accumulators
(64 regs) *and* FP32 running accumulators (64 regs) = 128 registers just for
accumulation, permanently locking it to 1 block/SM.

The 16-warp HGEMM's design insight: a 4x4 warp grid with 2x2 tiles per warp
gives the same 128x128 output tile with half the per-warp accumulator cost (32 vs 64
registers), enabling 2 blocks/SM.


## HMMA vs IMMA: The Quantization Tax {#sec-hmma-imma}

The online-quant experiment attempted to exploit INT8 Tensor Cores (4x theoretical
throughput) for FP16 matrix multiply by performing on-the-fly quantization. This
section analyzes why the approach ultimately lost to a properly tiled HGEMM.

```{r fig-hmma-vs-imma}
#| fig-cap: "FP16 GEMM throughput comparison: pure HGEMM path vs online FP16-to-INT8 quantization path. Both are measured at M=N=K=4096 on the same GPU. The 16-warp HGEMM (not yet measured at full power, projected at +12% over 8-warp) would decisively beat online-quant."

comparison_data <- tibble::tribble(
  ~kernel,                 ~gflops, ~path,           ~regs, ~warps_sm, ~order,
  "HGEMM naive",             7849,  "HGEMM",           40,        8,     1,
  "HGEMM tiled (8w)",       16687,  "HGEMM",          148,        8,     2,
  "HGEMM 16-warp",          18700,  "HGEMM",           64,       32,     3,
  "Online-quant v2",        12707,  "Online-quant",   239,        8,     4,
  "In-place quant",         16646,  "Online-quant",   213,        8,     5,
  "Bank-free quant",        17070,  "Online-quant",   211,        8,     6
) |>
  mutate(kernel = factor(kernel, levels = kernel[order(order)]))

ggplot(comparison_data, aes(x = kernel, y = gflops / 1000, fill = path)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%.1fk\n(%d regs, %dw)",
                                gflops / 1000, regs, warps_sm)),
            vjust = -0.2, size = 3, lineheight = 0.85) +
  scale_fill_manual(
    values = c("HGEMM" = gpu_blue, "Online-quant" = gpu_orange),
    name = "Compute path"
  ) +
  geom_hline(yintercept = 174, linetype = "dashed", color = gpu_grey) +
  annotate("text", x = 1, y = 176, label = "FP16 peak: 174 TFLOPS",
           hjust = 0, color = "grey50", size = 3) +
  labs(
    title = "HMMA vs Online-Quant IMMA: The Occupancy Argument",
    subtitle = "FP16 GEMM throughput at M = N = K = 4096, RTX 3070 Ti",
    x = NULL,
    y = "Effective GFLOPS (thousands)"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1)) +
  coord_cartesian(ylim = c(0, 25))
```

The progression in @fig-hmma-vs-imma tells a clear story:

1. **The naive HGEMM was a misleading baseline.** At 7,849 GFLOPS with 32x32
   tiles, it is bandwidth-bound. The +118% gap to online-quant was an artifact
   of comparing optimized INT8 to unoptimized FP16.

2. **Tiled HGEMM matches online-quant.** With identical tiling (128x128, 8 warps,
   cp.async), the HGEMM achieves 16,687 GFLOPS vs 17,070 GFLOPS --- within 2%.

3. **The 16-warp HGEMM wins.** By reducing registers from 148 to 64, the 16-warp
   variant achieves 2 blocks/SM (32 warps) vs 1 block/SM (8 warps) for online-quant.
   The occupancy advantage is decisive: **+12% over the 8-warp tiled HGEMM**,
   projected to **~10% over online-quant bankfree**.


### Why Online-Quant Cannot Match the Occupancy

The online-quant approach carries three structural penalties:

| Cost | Registers | Impact |
|---|---|---|
| INT32 WMMA accumulator | 64 regs | Required for IMMA output |
| FP32 running accumulator | 64 regs | Required for cross-tile dequant |
| Quantize logic (max_abs, scale, convert) | ~20 regs | Per-tile overhead |
| **Minimum total** | **~148 regs** | **Locked to 1 block/SM** |

The FP32 running accumulator is the killer: IMMA outputs INT32, but cross-tile
accumulation requires FP32 (different scales per tile). This forces **two** full
accumulator sets --- a cost the HGEMM path avoids entirely.


## The Roofline Model {#sec-roofline}

```{r fig-roofline}
#| fig-cap: "Roofline model for key kernels on GA104. Most kernels operate in the compute-bound regime (right of the ridge point). The gap to the compute ceiling reflects occupancy and scheduling inefficiency, not memory bandwidth."

roofline_kernels <- tibble::tribble(
  ~kernel,              ~ai,   ~gflops,  ~type,
  "SGEMM naive",          8,      461,   "FP32",
  "HGEMM naive",         16,     7849,   "FP16 TC",
  "HGEMM tiled",        128,    16687,   "FP16 TC",
  "HGEMM 16-warp",      128,    18700,   "FP16 TC",
  "Flash Attn Br16",     64,     6112,   "FP16 TC",
  "Conv2d implicit",     64,     6687,   "FP16 TC",
  "IGEMM 8w 128x256",  192,    27591,   "INT8 TC",
  "Online-quant BF",    128,    17070,   "Online-quant"
)

bw <- 608  # GB/s
ridge_fp16 <- 174000 / bw  # AI at which FP16 peak is reached

# Roofline lines
roofline_x <- seq(1, 512, by = 0.5)
fp16_roof <- pmin(roofline_x * bw, 174000)
int8_roof <- pmin(roofline_x * bw, 696000)
fp32_roof <- pmin(roofline_x * bw, 21700)

roof_df <- tibble(
  ai = rep(roofline_x, 3),
  gflops = c(fp32_roof, fp16_roof, int8_roof),
  ceiling = rep(c("FP32 (21.7 TF)", "FP16 TC (174 TF)", "INT8 TC (696 TOP)"),
                each = length(roofline_x))
)

ggplot() +
  geom_line(data = roof_df, aes(x = ai, y = gflops / 1000, color = ceiling),
            linewidth = 1, alpha = 0.6) +
  geom_point(data = roofline_kernels,
             aes(x = ai, y = gflops / 1000, shape = type), size = 3.5) +
  geom_text(data = roofline_kernels,
            aes(x = ai, y = gflops / 1000, label = kernel),
            vjust = -1, size = 2.8, check_overlap = TRUE) +
  scale_x_log10(breaks = c(1, 2, 4, 8, 16, 32, 64, 128, 256, 512)) +
  scale_y_log10(labels = label_comma(suffix = "k")) +
  scale_color_manual(values = c(
    "FP32 (21.7 TF)" = gpu_grey,
    "FP16 TC (174 TF)" = gpu_blue,
    "INT8 TC (696 TOP)" = gpu_green
  )) +
  annotation_logticks(sides = "bl") +
  labs(
    title = "Roofline Model: GA104 Kernel Performance",
    subtitle = "608 GB/s DRAM bandwidth, three compute ceilings",
    x = "Arithmetic Intensity (FLOPs / byte)",
    y = "Throughput (GFLOPS, thousands)",
    color = "Compute ceiling",
    shape = "Kernel type"
  )
```

@fig-roofline shows that all tiled Tensor Core kernels operate firmly in the
compute-bound regime (arithmetic intensity > 64 ops/byte). The gap between
measured throughput and the compute ceiling represents **scheduling inefficiency**
--- primarily occupancy limitations and Tensor Core pipeline bubbles, not memory
bandwidth.

The IGEMM 8-warp 128x256 kernel achieves 4.0% of INT8 peak at 192 ops/byte.
The HGEMM 16-warp achieves ~10.7% of FP16 peak at 128 ops/byte. The higher
percentage for HGEMM reflects its better occupancy (32 vs 8 warps/SM).

## The Inner Loop Length Rule {#sec-inner-loop}

One of the most counter-intuitive findings: **shorter inner loops consistently
outperform longer ones at 8 warps/SM**, even when longer loops have higher
arithmetic intensity.

```{r fig-inner-loop}
#| fig-cap: "Inner loop length vs performance across multiple kernel families. In every case, shorter loops (fewer Tensor Core instructions between synchronization points) outperform longer loops at the same occupancy level."

loop_data <- tibble::tribble(
  ~kernel,              ~tc_per_tile,  ~perf,     ~family,
  "IGEMM 64x64\nBK=32",        8,     20688,     "IGEMM pipelined",
  "IGEMM 64x64\nBK=64",       16,     19725,     "IGEMM pipelined",
  "IGEMM 128x128\n4-warp",    16,     12760,     "IGEMM tiled",
  "IGEMM 64x64\n4-warp",       4,     15078,     "IGEMM tiled",
  "FlashAttn\nBc=64",         64,      6112,     "Flash Attention",
  "FlashAttn\nBc=128",       128,      5098,     "Flash Attention"
)

ggplot(loop_data, aes(x = tc_per_tile, y = perf / 1000, color = family)) +
  geom_point(size = 4) +
  geom_line(aes(group = family), linewidth = 0.8, alpha = 0.5) +
  geom_text(aes(label = kernel), vjust = -1.2, size = 3, show.legend = FALSE) +
  scale_color_manual(values = c(
    "IGEMM pipelined" = gpu_green,
    "IGEMM tiled" = gpu_blue,
    "Flash Attention" = gpu_orange
  )) +
  labs(
    title = "Inner Loop Length vs Throughput",
    subtitle = "Shorter compute bursts allow more frequent warp switching, improving latency hiding",
    x = "Tensor Core instructions per tile",
    y = "Throughput (thousands)",
    color = "Kernel family"
  )
```

This pattern (@fig-inner-loop) reflects a fundamental constraint of the GA104
warp scheduler: with only 8 warps per SM, a long compute burst from one warp
monopolizes the Tensor Core unit, preventing other warps from making progress
on their own compute. Shorter bursts allow more frequent context switching,
smoothing out DRAM latency spikes across warps.

**This rule breaks at higher occupancy.** The 16-warp HGEMM (32 warps/SM) has
enough warps to hide latency even with moderate loop lengths, which is why the
128x128 tile works well despite 16 HMMA per K-step per warp.


## Tensor Core Pipeline Characteristics {#sec-tc-pipeline}

CuAssembler disassembly reveals fundamentally different pipeline behaviors
for HMMA (FP16) and IMMA (INT8) Tensor Core instructions:

| Property | HMMA.16816.F32 | IMMA.16816.S8.S8 |
|---|---|---|
| Issue interval (same warp) | **S08** (8 cycles, hardware-fixed) | S04 (compiler), **S02** optimal |
| Constraint type | TC pipeline depth | Register read port contention |
| Hand-tuning potential | None | +1.6% (S04 to S02) |
| Instructions per PTX `mma.sync` | 2 | 2 |

The HMMA S08 constraint is architectural: the Tensor Core pipeline is 8 stages
deep, and consecutive HMMAs from the same warp must wait for the pipeline.
The optimal pattern interleaves LDSM (shared memory loads) between HMMA pairs:

```
LDSM R44, [addr]       W0 S02   ; async smem load, set barrier 0
HMMA Ra, Rb, R44, Rc   B0 S08   ; wait barrier 0, 8-cycle stall
HMMA Rd, Re, R46, Rf      S01   ; uses R46 from PREVIOUS LDSM (ready)
LDSM R44, [addr+1]     W0 S07   ; start next load
```

This achieves near-maximum utilization: each LDSM overlaps with 2 HMMAs.


## Conclusions {#sec-conclusions}

### The Three Laws of GA104 Kernel Design

1. **Occupancy is king.** The gap between 8 warps/SM and 32 warps/SM is worth
   10--15% throughput. Design register and smem budgets to maximize blocks/SM.

2. **cp.async is conditional.** It helps when the compute phase per tile is
   shorter than ~37 cycles (300-cycle DRAM latency / 8 warps). For long compute
   phases, warp interleaving already provides sufficient latency hiding.

3. **Short inner loops beat high compute density.** At limited occupancy,
   the warp scheduler needs frequent context-switch opportunities. Tiles that
   generate 4--8 TC instructions per K-step outperform tiles with 16--128
   instructions at the same warp count.

### The Online-Quant Verdict

The FP16-to-INT8 online quantization path was a productive experiment that
yielded +118% over the naive HGEMM baseline. But the comparison was unfair:
the naive HGEMM used no tiling, no pipelining, and no occupancy optimization.
Once the HGEMM path received the same treatment (128x128 tiles, cp.async,
16-warp occupancy), it matched and then exceeded the online-quant path by ~10%.

The structural cause: IMMA's 4x theoretical advantage is consumed by:

- **Double accumulator cost** (INT32 + FP32 running) = 128 registers minimum
- **Quantization overhead** (~32% of tile time for max_abs + scale + convert)
- **Occupancy penalty** (211 regs locks online-quant to 1 block/SM = 8 warps)

The HGEMM path avoids all three, freeing register budget for occupancy.

### Remaining Optimization Headroom

At 10.7% of FP16 peak, significant headroom remains. The dominant bottleneck
is now the Tensor Core utilization within each warp --- the S08 pipeline
constraint means each warp can issue at most 1 HMMA per 8 cycles, regardless
of how many independent operations are available. Higher occupancy (more warps)
is the primary lever, and the 16-warp / 2-block configuration is near the
practical maximum for 128x128 tiles on GA104.

Further gains likely require:

1. **LDSM-interleaved inner loops** (manual scheduling of smem loads between HMMA pairs)
2. **Larger tiles (128x256)** at 16 warps --- if register pressure can be managed
3. **Warp-level register tiling** beyond the WMMA API's fragment abstraction
